\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-30 papers/jfp/7-experience.tex
Finish S7.2: Hadrian

\todo{AM}{Finish this section.}

The build system used to compile GHC, known as Hadrian, has continued
development. The use of dynamic dependencies has made the build much easier, and
it's now more maintainable. It's now merged into the GHC repo and tested as
standard by the CI. It's current state is ... In particular, it takes N minutes
to build with \Shake, vs M minutes to build with \Make. The \Shake version has
attracted more contributors, and been easier to modify, etc. The code is N
lines.

The design we originally outlined remains the one in use. There has been a huge
amount of engineering work to make the tests pass (many of the details encoded
in the \Make build system were incidental -- but some were essential).

One of the benefits of using Shake is that we have access to high quality
profiling information, allowing us to compute critical paths and other metrics,
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for a tour of the metrics.
This information has shown us that GHC is slow to build (40m on an 8 CPU machine
on Windows), that more CPUs would not help (on unlimited CPUs it would take at
least 37m), and that a handful of steps (two Haskell compiles, some calls to
\cmd{configure}) are responsible for a significant amount of that time (at least
10m).

>>>>

\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
developed to replace a \Make-based build system and solve multiple scalability
and maintainability challenges. As discussed in detail
by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
of two key \Make's shortcomings: (i)~poor abstraction facilities of makefiles,
i.e. the need to program in a single namespace of mutable string variables, and
(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
\Hadrian benefits both from \Shake's features and from the host language
Haskell, making the new GHC build system easier to understand and maintain.
Interestingly, although \Shake is not a self-tracking build
system~(\S\ref{sec-tracking-aspects}), \Hadrian implements a little
domain-specific language for constructing build command lines, and then tracks
command lines by treating them as a type of values -- an example of
\emph{partial self-tracking} made possible by \Shake's support for value
polymorphism~(\S\ref{sec-polymorphism}).

The development of \Cloud \Shake allows GHC developers to take advantage of
sharing build results between builds, which is particularly promising in the
context of GHC's continuous integration (CI) infrastructure responsible for
building and testing every GHC commit. Building GHC~8.8 from scratch with
default settings takes around 1 hour on Windows using both \Hadrian and the
original \Make-based build system. This includes building the compiler itself,
29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
only a small part of the whole codebase, making the cloud-build functionality
highly desirable. In particular, if a commit does not touch any GHC source files
at all (e.g. only modifies the documentation), \Hadrian can build GHC in under
3~minutes, by simply creating symbolic links to the previously built results
stored in the \Cloud~\Shake cache. Note that a small number of GHC build rules
cannot be cached since they mutate the bundled package database (e.g. to
register a new library); \Cloud~\Shake provides a way to manually label such
build rules, thereby excluding them from caching.

One of the benefits of using \Shake is that we have access to high quality build
profiling information, allowing us to compute critical paths and other metrics,
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
profiling features. This information has shown us that GHC is slow to build
(1~hour on an 4~CPU Windows machine), that more CPUs would not help (on
unlimited CPUs the speed up would be less than 10\%), and that a handful of
build tasks (two anomalously slow Haskell compilations, and some calls to slow
single-threaded \cmd{configure}) are responsible for a significant amount of
build time (at least 10~minutes).

% Benchmarking:
% \begin{itemize}
%   \item Building \Hadrian: 55s. For comparison, building the Cabal library: 4m.
%   \item Initial Windows tarball download: 45s.
%   \item Initial \cmd{boot} and \cmd{configure} on Windows: 2m 25s.
%   \item Full GHC build using \Hadrian with default settings:
%         \cmd{hadrian/build.bat -j8}: 57m. Subsequent zero build: 5s.
%   \item Full GHC build using \Hadrian with default settings in the cloud mode:
%         \cmd{hadrian/build.bat -j8 -}\cmd{-share=cache}: 58m.
%         Subsequent zero build: 5s.
%         Subsequent full rebuild after deleting the build directory: 5m 27s.
%   \item Full GHC build using \Make with default settings: \cmd{make -j8}: 60m.
%         Subsequent zero build: 18s.
% \end{itemize}

% The build system used to compile GHC, known as \Hadrian, has continued
% development. The use of dynamic dependencies has made the build much easier, and
% it's now more maintainable. It's now merged into the GHC repo and tested as
% standard by the CI. It's current state is ... In particular, it takes N minutes
% to build with \Shake, vs M minutes to build with \Make. The \Shake version has
% attracted more contributors, and been easier to modify, etc. The code is N
% lines.

% The design we originally outlined remains the one in use. There has been a huge
% amount of engineering work to make the tests pass (many of the details encoded
% in the \Make build system were incidental -- but some were essential).
